---
title: "Scotland Covid Impact R Model (SCIRe model)"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Model Outline and Outputs

The main outputs from the model are:  
  
* Spreadsheets detailing our current estimates of cases with symptoms, total hospitalisation, cases requiring ventilation, deaths, and equipment and medicine required by day and by week.  
* Graphs showing the number of people needing hospital beds and intensive care beds by day and by week.

## Requirements for running the model

Functions are stored in multiple R files which should be saved in /Code/functions. Required packages are called within the functions. All functions can then be sourced using the header.R file:

```{r}
#source("x.R") #if you have files saved locally 
source("x.R") #if you are accessing files through x
```

Data and assumptions files should be saved in Data/, and templates for the output should be saved in Output templates/. The following folders need to be set up to allow the model to save the output spreadsheets and charts in:  
  
* Outputs/Current estimates spreadsheets,  
* Outputs/curves,  
* Outputs/raw.

## Assumptions used

The model uses assumptions from the file "Model Assumptions.xlsx". The assumptions used in the "No intervention" scenario for people aged 40 to 49 can be seen by running the follwing code:

```{r}
library(knitr)
"Scenario setup.xlsx" %>% 
    get_filepath("Data") %>% 
    get_scenarios %>% 
    mutate(data = ID %>%
             map(prepare_assumptions, Age_group)) %>% # Age_group, Vulnerable, In_care, SIMD for full model
    unnest %>% 
  filter(Age_group==40,
         ID==0) %>%
  select(Scenario_name, Age_group, `Curve file`,`Curve method`, Assumption, Value) %>%
  kable(caption="Assumptions used for people aged 40-49 in the 'No intervention' scenario")
```

## Running the model
The assumptions are run through the function "run_model" to calculate infection and mortality rates for each demographic at each severity level of the disease. Severity levels, along with the definitions used in the model, are:  
  
* Mild (asymptomatic),  
* Moderate (symptomatic, but will not require hospitalisation),  
* Serious (will require oxygen in hospital, but not ventilation),  
* Severe (will require non-invasive ventilation, but not invasive ventilation),  
* Critical (will require invasive ventilation).  
  
Demographics can be broken down by age group, SIMD level, vulnerability, and caring status by including these as inputs in the function "run_model". As a demonstration, here we will break down by age group and look at the rates for people aged 40-49 in the "No intervention" scenario.

```{r}
run_model(Age_group) %>% #run_model(Age_group, SIMD, Vulnerable, In_care) for full model
  filter(ID==0,
         Age_group==40) %>%
  select(-ID) %>%
  kable(caption="Infection and mortality rates used for people aged 40-49 in the 'No intervention' scenario")
```

Applying these rates to the population will give an estimate of the number of people who will be infected or die from the disease. To estimate the numbers at each point in time, we first combine these rates with a timeline.

There is a curve file named in the assumptions table above. For each scenario we use a different curve to estimate the proportion of infections at each point in time. 

```{r}
curve = get_curve() %>%
  filter(ID==0,
         Age_group==40)

curve %>% ggplot(aes(Date,proportion_in_period)) +
  geom_line(size=1.1) +
  theme_light() +
  ggtitle("Proportion of cases over time, in people aged 40-49, in the 'No intervention' scenario") +
  scale_color_brewer(palette = "Dark2") +
  theme(axis.title.y = element_blank())

```

We combine this curve with an estimated timeline for each infection, which differs based on demographic and severity level of the infection.

The timeline from the start date of infection for people aged 40-49 with a severe illness is shown below.

```{r}
get_timeline(length_of_infection=13,
             date_of_admission_from_infection_start=9,
             treatment_recovery_factor=1) %>%
  filter(In_care=="No",Age_group==40,severity=="Severe",
         Status %in% c("Asymptomatic", "Symptomatic", "Recovering", "Recovered"),
         Stat == "Proportion_at") %>%
  rename("Days from infection start"=date_from_infection) %>%
  ggplot(aes(`Days from infection start`, proportion, fill=Status)) +
  geom_bar(position = 'stack', width=1, stat="identity") +
    ggtitle("Proportion of people at different stages of disease from infection start date") +
    theme(axis.title.y = element_blank()) +
    scale_fill_brewer(palette = "Set2")
```

The chart above excludes treatment statuses, which are a subset of the Symptomatic status. The Dead status is also excluded as at this point in the model there is no determination as to what proportion live and die, and so including the Dead status alongside Recovering and Recovered would lead to double counting. This table includes all statuses for people aged 40-49 with a severe illness:

```{r}
get_timeline(length_of_infection=13,
             date_of_admission_from_infection_start=9,
             treatment_recovery_factor=1) %>%
  filter(In_care=="No",Age_group==40,severity=="Severe") %>%
  select(date_from_infection, Stat, Status, proportion) %>% 
  spread(date_from_infection, proportion, fill=0)
```

By the combining the curve, the timelines, and infection and mortality rates, we can estimate the proportion of people infected, requiring treatment, recovering, recovered, and dead at each point in time. We calculate the proprotion of people moving into each category (Proportion_starting) and in each category at each point in time (Proportion_at).

```{r}
model=run_model(Age_group)

proportions = run_timeline(Age_group) %>% 
  get_proportions(model) 

proportions %>% 
  filter(ID==0, Age_group==40, Date<ymd("2020-09-01")) %>%
  group_by(Scenario_name,Date,Status) %>%
  summarise(Proportion_starting=sum(Proportion_starting)) %>%
  ggplot(aes(Date,Proportion_starting,colour=Status,group=Status)) +
      geom_line(size=1.1) +
      theme_light() +
      ggtitle("Proportion_starting, for people aged 40-49 in the 'No intervention' scenario") +
      scale_color_brewer(palette = "Dark2") +
      theme(axis.title.y = element_blank())

proportions %>% 
  filter(ID==0, Age_group==40, Date<ymd("2020-09-01")) %>%
  group_by(Scenario_name,Date,Status) %>%
  summarise(Proportion_at=max(Proportion_at)) %>%
  ggplot(aes(Date,Proportion_at,colour=Status,group=Status)) +
      geom_line(size=1.1) +
      theme_light() +
      ggtitle("Proportion_at, for people aged 40-49 in the 'No intervention' scenario") +
      scale_color_brewer(palette = "Dark2") +
      theme(axis.title.y = element_blank())
```

Next we apply these proportions to the population of each demographic in order to estimate of the number of people infected, requiring treatment, recovering, recovered, and number of fatalities for each demographic in each intervention scenario. 

```{r}
get_population(Age_group) %>% #get_population(Age_group, SIMD, Vulnerable, In_care)
  mutate(Age_group = recode(Age_group,
                            `0`="0 to 9", 
                            `10`="10 to 19",
                            `20`="20 to 29",
                            `30`="30 to 39",
                            `40`="40 to 49",
                            `50`="50 to 59",
                            `60`="60 to 69",
                            `70`="70 to 79",
                            `80`="80 to 89",
                            `90`="90+")) %>%
  kable(caption="Population of people in Scotland, by age group")
```

Number_starting gives the number of new people in each category at each point in time, and Number_at gives the total number of people in each category at each point in time. We also set the date here.
```{r}
population = get_population(Age_group)

scotland_estimates = proportions %>% 
    get_totals(population)
scotland_estimates %>% 
  filter(ID==0, Age_group==40, Date<ymd("2020-09-01")) %>%
  group_by(Scenario_name,Date,Status) %>%
  summarise(Number_starting=sum(Number_starting)) %>%
  ggplot(aes(Date,Number_starting,colour=Status,group=Status)) +
      geom_line(size=1.1) +
      theme_light() +
      ggtitle("Number_starting, for people aged 40-49 in the 'No intervention' scenario") +
      scale_color_brewer(palette = "Dark2") +
      theme(axis.title.y = element_blank())

scotland_estimates %>% 
  filter(ID==0, Age_group==40, Date<ymd("2020-09-01")) %>%
  group_by(Scenario_name,Date,Status) %>%
  summarise(Number_at=max(Number_at)) %>%
  ggplot(aes(Date,Number_at,colour=Status,group=Status)) +
      geom_line(size=1.1) +
      theme_light() +
      ggtitle("Number_at, for people aged 40-49 in the 'No intervention' scenario") +
      scale_color_brewer(palette = "Dark2") +
      theme(axis.title.y = element_blank())
```


## Outputs
These numbers are used to populate the output spreadsheets and charts by running the line of code below. Spreadsheets are saved into the /Outputs/Current estimates spreadsheets/ folder, and charts are saved into the /Outputs/curves/ folder. We use the charts to update our slide pack daily.
```{r eval=FALSE}
run_daily_update()

create_excel_output(ymd("2020-09-27"))
create_slidepack_output(week_min = 20, week_max = 35)
 
```
